---
title: "Project 2"
author: "OilGasDataAnalyst"
date: "October 20, 2015"
output: html_document
---

###Title:
NOAA Storm Database Severe Weather Analysis

###Synopsis:
The NOAA has a database on severe weather. This database goes back to the 1950's however was standarized in 1996. The data covers major weather events and their impact on human lives, economic impact along with location data. By analyizing the data one can gain an understanding of the impact the different weather events pose. This data can then be used by government and private individuals to prepare for future events as well as understand what has the greatest impacts. This analysis plans to understand what type of event have the greatest impact on human population health and economic impact.

###Data Processing
Data processing section will walk through the steps taken to process & create the plots used in the data analysis.

Start with loading the librarys
```{r}
library(plyr)
```

Now we need to read in the data.
```{r, cache=TRUE}
StormData <- read.csv("repdata-data-StormData.csv.bz2")
head(StormData)
```

After analyizing the data structure and reading the information from http://www.ncdc.noaa.gov/stormevents/details.jsp as found on the course forums, one can say we should only investigate incidents after 1996. Otherwise tornados will have an unusally high amount of injuries & damage. This still gives us 20 years of data to work with, time permitting we can look back at the rest of the pre-1996 data and see if its relatively inline with the post 1996 data.

So we need to extract the data after 1996. So first convert the dates and check that everything has converted properly.
```{r,cache=TRUE}
StormData$BGN_DATE <- as.Date(StormData$BGN_DATE, format = "%m/%d/%Y %H:%M:%S")
class(StormData$BGN_DATE)
sum(is.na(StormData$BGN_DATE))
```

Now we want to subset the data to a working range of data that we want to deal with.
```{r,cache=TRUE}
WorkingData <- subset(StormData, BGN_DATE >= "1996-01-01")
```

First we want to investigate how harmful to population health, i.e fatalities & injuries.
```{r, cache=TRUE}
HumanFatal <- aggregate(FATALITIES ~ EVTYPE,WorkingData,sum)
HumanInjury <- aggregate(INJURIES ~ EVTYPE,WorkingData,sum)
HumanImpact <- merge(HumanInjury, HumanFatal, by = "EVTYPE")
```

We now have a subset of data of the event types & number of injuries & fatalities that each of the event types have caused. Taking a look at the data `view(HumanImpact)` within RStudio we can see that a lot of the values have 0 for either fatalities or injuries. So lets go ahead and subset the data set down some more.
```{r, cache=TRUE}
HumanImpact <- subset(HumanImpact, (INJURIES > 0 & FATALITIES > 0))
```

We now want to get a combined total of injuries & fatalities for sorting on
```{r, cache=TRUE}
HumanImpact$Combined <- HumanImpact$INJURIES + HumanImpact$FATALITIES
HumanImpact <- HumanImpact[order(-HumanImpact$Combined),]
```

Taking a look at the data we can see clearly that Tornados have quite the impact!
```{r, cache=TRUE}
head(HumanImpact)
```

Lets plot out the data in a stacked bar chart, which is surprisingly hard! First we grab the top ten events and factor the event names.
```{r, cache=TRUE}
plotdata <- HumanImpact[1:10, 1:4]
plotdata$EVTYPE <- factor(plotdata$EVTYPE)
```

We then need to manipulate the data so that we can graph out as a bar chart, this needs to be a matrix, so first create a new data frame.
```{r, cache=TRUE}
Injury <- c(plotdata$INJURIES)
Death <- c(plotdata$FATALITIES)
HumanDF <- data.frame(Injury, Death)
```

Then transpose the dataframe with `t()` and apply the event names across so that the x-axis is properly labeled. The `HumanDF` dataframe will be used to plot out the data in the results section.
```{r, cache=TRUE}
HumanDF <- t(HumanDF)
colnames(HumanDF) <- plotdata$EVTYPE
```

Now we move onto the second question, regarding economic consequences. First we need to take a look at the data:
```{r, cache=TRUE}
head(WorkingData,1)
```

From this we can identify four columns of data that we need to utilize, `PROPDMG`, `PROPDMGEXP`,`CROPDMG` & `CROPDMGEXP` in addtion to `EVTYPE`. So lets subset out this data into our new working variable, lets call it `EconData`
```{r, cache=TRUE}
EconData = data.frame(factor(WorkingData$EVTYPE),WorkingData$PROPDMG,WorkingData$PROPDMGEXP,WorkingData$CROPDMG,WorkingData$CROPDMGEXP)
colnames(EconData) <- c("EVTYPE","PROPDMG","PROPDMGEXP","CROPDMG","CROPDMGEXP")
head(EconData)
```

We can see lots of zero values, so lets filter these out.
```{r, cache=TRUE}
EconData <- subset(EconData, (PROPDMG > 0 & CROPDMG > 0))
```

Now according to the Storm Data Export document the EXP columns will be a multiplyer (if any) to apply to the damage amount, so lets multiply these across using the amazing `mapvalues` function within plyr. Also convert them to numeric.
```{r, cache=TRUE}
EconData$PROPDMGEXP <- as.character(EconData$PROPDMGEXP)
EconData$CROPDMGEXP <- as.character(EconData$CROPDMGEXP)

EconData$PROPDMGEXP <- mapvalues(EconData$PROPDMGEXP,c("K","M","B"),c(1000,1000000,1000000000))
EconData$CROPDMGEXP <- mapvalues(EconData$CROPDMGEXP,c("K","M","B"),c(1000,1000000,1000000000))

EconData$PROPDMGEXP <- sapply(EconData$PROPDMGEXP,as.numeric)
EconData$CROPDMGEXP <- sapply(EconData$CROPDMGEXP,as.numeric)
```

Now find the total value of damage by multiplying the mulitplyer `PROPDMGEXP` & `CROPDMGEXP` by the damage values.
```{r, cache=TRUE}
EconData$PROPDMG <- EconData$PROPDMG * EconData$PROPDMGEXP
EconData$CROPDMG <- EconData$CROPDMG * EconData$CROPDMGEXP
head(EconData)
```

We will now repeate the same process as preping the injury data.

1. Convert `EVTYPE` to a factor to remove the levels

2. Aggregate the damages by `EVTYPE`

3. Add up these damages to sort by largest

4. Sort code

5. Grab the top 10 items for plotting 
```{r, cache=TRUE}
EconData$EVTYPE <- factor(EconData$EVTYPE)
EconPlot <- aggregate(cbind(PROPDMG,CROPDMG) ~ EVTYPE, EconData, sum)
EconPlot$TotalDMG <- EconPlot$PROPDMG + EconPlot$CROPDMG
EconPlot <- EconPlot[order(-EconPlot$TotalDMG),]
EconPlot <- EconPlot[1:10, ]
```

Then we make the dataframe consisting of the `PROPDMG` & `CROPDMG`, transpose the dataframe and the add the names to the columns so they appear on the bar chart.
```{r, cache=TRUE}
EconDF <- data.frame(EconPlot$PROPDMG,EconPlot$CROPDMG)
EconDF <- t(EconDF)
colnames(EconDF) <- EconPlot$EVTYPE
```

###Results

1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

Plotting out the data of the event types we can see tornados are the biggest threat in respect to population health.
```{r, cache=TRUE}
par(mar=c(7,5,1,1))
barplot(as.matrix(HumanDF),cex.names = .6,col=c("black","red"),las = 2,ylab = "Count of Injuries/Fatalities", legend = c("Injuries","Fatalities"),mgp=c(3,.5,0))
```

Taking a look at the data it is easy to see that tornados on their own are the biggest threat. However this in turn is simply a bad storm which coul consist of lightning, flash flooding, thunderstorm winds which also comprise the top 10 items. So in general storms appear to be the biggest threat to life. These are often difficult to prepare for as they are difficult to predict & planning for a tornado is minimal. Most regions in the "tornado alley" are well prepared and have warning systems in place for tornados.

As far as planning excessive heat is one area which could actually be prevented and human impact cut back on. Limits to activity outdoors and required breaks for workers outdoors could lower this number. More analysis into the data specifics is required.

2. Across the United States, which types of events have the greatest economic consequences?

```{r, cache=TRUE}
par(mar=c(7,7,1,1))
barplot(as.matrix(EconDF),cex.names = .6,col=c("black","red"),las = 2,ylab = "Nominal Sum of Damages USD$", legend = c("Property Damage","Crop Damage"),mgp=c(5,.5,0))
```

For economic consequences flooding is the clear "winner" here. Flooding far exceeds the next two items which in turn are often a cause of flooding themselves! As far as preparing for flooding it really boils down to investigating flood zones and preventing/placing restricions on building in those areas. Due to human nature living in a flood zone will be a way of life in general, however specific precautions can be taken to prevent damage where possible.

###Conclusions
The above report could be broken down some more. The specific areas that could be addressed are:

1. Aggregating the `EVTYPE` even further, there are instances of "HURRICANE/TYPHOON" & "HURRICANE" which could/should be combined. This is debateable as I filtered the dataset to post 1996 data which should have corrected this, taking a look at the unique values we can see that HURRICANE is standalone, which could be corrected and then combined. Others could be combined as well (WINTER STORM & HEAVY SNOW), however the decision to filter to post 1996 data was to prevent this issue and gain a greater understanding of what is causing the greatest harm.

```{r, cache=TRUE}
unique(EconData$EVTYPE)
```

2. Investigation into the `REMARKS` field could yield greater information regarding specific events, such as the EXCESSIVE HEAT item. During my inital data exploration there appears to be a wealth of information reagarding the hows & whys as to the injuries & deaths. This data itself could be explored as to the actual cause of death per `EVTYPE` and that would give a better idea as to where to focus efforts regarding planning for these events.