---
title: "Project 2"
author: "OilGasDataAnalyst"
date: "October 20, 2015"
output: html_document
---

###Title:
NOAA Storm Database Severe Weather Analysis

###Synopsis:
Immediately after the title, there should be a synopsis which describes and summarizes your analysis in at most 10 complete sentences.

###Data Processing
There should be a section titled Data Processing which describes (in words and code) how the data were loaded into R and processed for analysis. In particular, your analysis must start from the raw CSV file containing the data. You cannot do any preprocessing outside the document. If preprocessing is time-consuming you may consider using the cache = TRUE option for certain code chunks.

First we need to read in the data.
```{r, cache=TRUE}
StormData <- read.csv("repdata-data-StormData.csv.bz2")
```

After analyizing the data structure and reading the information from http://www.ncdc.noaa.gov/stormevents/details.jsp as found on the course forums, once can say we should only investigate incidents after 1996. Otherwise tornados will have an unusally high amount of injuries & damage. This still gives us 20 years of data to work with, time permitting we can look back at the rest of the pre-1996 data and see if its relatively inline with the post 1996 data.

So we need to extract the data after 1996.
```{r,cache=TRUE}

```

###Results

Your data analysis must address the following questions:

####Important Note!:

From http://www.ncdc.noaa.gov/stormevents/details.jsp

Event Types Available: 
Add more info about event types here. Link to collections page/tab when referencing data collection source. 

1. Tornado: From 1950 through 1954, only tornado events were recorded.

2. Tornado, Thunderstorm Wind and Hail: From 1955 through 1992, only tornado, thunderstorm wind and hail events were keyed from the paper publications into digital data. From 1993 to 1995, only tornado, thunderstorm wind and hail events have been extracted from the Unformatted Text Files.

3. All Event Types (48 from Directive 10-1605): From 1996 to present, 48 event types are recorded as defined in NWS Directive 10-1605. 

With this in mind will analyize data from 1996 onwards.


1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

2. Across the United States, which types of events have the greatest economic consequences?

Consider writing your report as if it were to be read by a government or municipal manager who might be responsible for preparing for severe weather events and will need to prioritize resources for different types of events. However, there is no need to make any specific recommendations in your report.

There should be a section titled Results in which your results are presented.

You may have other sections in your analysis, but Data Processing and Results are required.

The analysis document must have at least one figure containing a plot.

Your analyis must have no more than three figures. Figures may have multiple plots in them (i.e. panel plots), but there cannot be more than three figures total.

You must show all your code for the work in your analysis document. This may make the document a bit verbose, but that is okay. In general, you should ensure that echo = TRUE for every code chunk (this is the default setting in knitr).
